---
title: "multimod to GMPL"
author: "multimod Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{multimod to GMPL}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE to run during build
)
```

## Introduction

This vignette demonstrates the complete workflow for converting an energyRt model to GMPL (GNU MathProg Language, as subset of AMPL) format and solving it with GLPK. The workflow includes:

1. Reading a GAMS model structure
2. Converting to multimod format
3. Connecting to energyRt scenario data (stored as Arrow/Parquet)
4. Writing GMPL model and data files
5. Solving with GLPK using glpkAPI

We'll use the UTOPIA base scenario from energyRt as our example.

## GMPL-Specific Functions and Behavior

### Key Functions

#### 1. `write_gmpl(model, file, include_solve, export_vars, use_table_output)`

Writes the model structure to a `.mod` file.

**Parameters:**

- `include_solve = TRUE`: Adds solve statement, objective, and output code
- `export_vars`: Controls which variables to export to CSV
  - `NULL` (default): Export all variables to individual CSVs
  - `c("var1", "var2")`: Export only specified variables
  - `FALSE` or `character(0)`: No custom CSV export
- `use_table_output`: Adds GMPL table statement for output
  - `TRUE`: Exports all variables using table statement to `table_output.csv`
  - `FALSE` (default): No table output
- Both `export_vars` and `use_table_output` can be used together

**Example:**
```r
# Export only key variables
write_gmpl(model, "model.mod", export_vars = c("vTechCap", "vObjective"))

# Use GMPL table output instead of custom CSVs
write_gmpl(model, "model.mod", export_vars = FALSE, use_table_output = TRUE)
```

#### 2. `write_gmpl_data(model, file)`

Writes the data section to a `.dat` file with lazy loading from Arrow/Parquet files.

**GMPL-specific formatting:**

- Empty sets: `set name := ;`
- Empty mappings: `set name := ;`
- Empty parameters: `param name default <defVal> := ;`
- Multi-dimensional mappings use comma-separated tuples: `elem1,elem2, elem1,elem3, ...`
- Parameters with default values: `param name default <defVal> := [idx1,idx2] val ...`
- Sparse parameters: Default value ensures missing combinations are handled correctly
- File ends with: `end;`

#### 3. `populate_sets_from_scenario(model, scenario)`

Extracts unique set members from energyRt scenario parameters.

- Scans all Arrow/Parquet files to find dimension values
- Populates `model$sets[[name]]$data` with character vectors
- Required before writing data file

#### 4. `link_scenario_data(model, scenario)`

Links energyRt parameter data to model without loading into memory.

- Distinguishes mappings (`@type == "map"`) from parameters (`@type == "numpar"`)
- Preserves `@defVal` for parameter default values
- Stores references to Arrow/Parquet files in `model$misc$path`
- Data is loaded only when `write_gmpl_data()` is called (lazy loading)

### Key GMPL Differences from GAMS

| Feature | GAMS | GMPL |
|---------|------|------|
| Case sensitivity | Case-insensitive | Case-sensitive |
| Tuple syntax | `(a, b, c)` or `a.b.c` | Commas: `a,b,c` |
| Index syntax | `[a,b,c]` or `[a b c]` | `[a,b,c]` only |
| Empty declarations | Optional | Required |
| Sparse data | Auto-handled | Needs default values |
| Output | `put` statements | `printf` statements |
| Data I/O | `$include`, GDX | `table` statement |

## Prerequisites

```{r libraries}
library(multimod)
```

## Step 1: Read GAMS Model Structure

First, we read the model structure from the GAMS file. This gives us the model skeleton (sets, parameters, variables, equations) without data.

```{r read-gams}
# Path to GAMS model
gams_file <- "C:/Users/admin/Documents/R/multimod/dev/scenarios/BASE_UTOPIA/script/gams_cbc/energyRt.gms"

# Read model structure
cat("Reading GAMS model...\n")
model_struct <- read_gams(gams_file)

# Inspect structure
cat("Model structure:\n")
cat("  Sets:", length(model_struct$sets), "\n")
cat("  Mappings:", length(model_struct$mappings), "\n")
cat("  Parameters:", length(model_struct$parameters), "\n")
cat("  Variables:", length(model_struct$variables), "\n")
cat("  Equations:", length(model_struct$equations), "\n")
```

## Step 2: Convert to Multimod Format

Convert the model structure to multimod's internal representation. This automatically generates index aliases for use in GMPL.

```{r convert-multimod}
cat("\nConverting to multimod...\n")
model <- as_multimod(model_struct)

# Check index aliases
cat("Index aliases generated:\n")
print(model$index_aliases)
```

The index aliases map set names to short variable names used in equation iterations:
- `tech` → `h`
- `region` → `r`
- `comm` → `c`
- `year` → `y`
- etc.

These aliases are used in GMPL's `for` loops and indexing expressions.

## Step 3: Connect to energyRt Scenario Data

Now we connect the model to actual data from an energyRt scenario. The data is stored in Arrow/Parquet format on disk.

```{r load-scenario}
# Load energyRt scenario
cat("\nLoading energyRt scenario...\n")
library(energyRt)
scen_file <- "C:/Users/admin/Documents/R/multimod/dev/scenarios/BASE_UTOPIA/scen.RData"
load(scen_file)

cat("Scenario class:", class(scen), "\n")
cat("Parameters in scenario:", length(scen@modInp@parameters), "\n")

# Set base path for lazy loading
model$base_path <- "C:/Users/admin/Documents/R/multimod/dev/scenarios/BASE_UTOPIA"
cat("Base path set to:", model$base_path, "\n")
```

### Import Scenario Data with Logging

We use the comprehensive `import_energyRt_data()` function that handles sets, parameters, mappings, and bounds parameters:

```{r import-data}
# Import all data with detailed logging
cat("\nImporting energyRt scenario data...\n")
model <- import_energyRt_data(model, scen, inMemory = TRUE, log_file = "tmp/import_log.csv")

# Check what was imported
cat("\nImport summary:\n")
cat("  Sets populated\n")
cat("  Parameters linked\n")
cat("  Mappings linked\n")  
cat("  Bounds split and linked\n")
```

### Inspect Bounds Mapping

Bounds parameters (e.g., `pTechAf` with type "bounds") are automatically split into separate Lo/Up parameters:

```{r bounds-inspect}
# Show bounds mapping
show_bounds_mapping(model)

# Check for any import issues
show_unmatched(model)

# Access detailed import log
log <- get_import_log(model)
cat("\nBounds processed:", length(log$bounds), "\n")
```

The `import_energyRt_data()` function:

1. **Populates sets** from scenario parameter dimensions
2. **Links parameters** (numeric data) with lazy loading support
3. **Links mappings** (set membership data)
4. **Splits bounds parameters** into separate Lo/Up parameters
   - Example: `pTechAf` (type="bounds") → `pTechAfLo` + `pTechAfUp`
   - Handles dimension aliases (e.g., `src`/`dst` for `region`)
   - Preserves defVal: `c(lo_default, up_default)`
   - Filters by "type" column: `lo/lower/min` vs `up/upper/max`


## Step 4: Write GMPL Files

Now we write the model and data files. The data will be loaded from disk automatically during writing.

```{r write-gmpl}
# Create output directory
output_dir <- "C:/Users/admin/Documents/R/multimod/tmp/utopia_gmpl"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Write model file (.mod) with solve and output statements
cat("\nWriting GMPL model file...\n")
mod_file <- file.path(output_dir, "utopia.mod")
write_gmpl(model, file = mod_file, include_solve = TRUE)

cat("Model file written:", mod_file, "\n")
cat("File size:", format(file.size(mod_file), big.mark = ","), "bytes\n")

# Write data file (.dat) - this triggers lazy loading
cat("\nWriting GMPL data file (loading data from disk)...\n")
dat_file <- file.path(output_dir, "utopia.dat")
write_gmpl_data(model, file = dat_file)

cat("Data file written:", dat_file, "\n")
cat("File size:", format(file.size(dat_file), big.mark = ","), "bytes\n")
```

### Output Control Options

You can control which variables are exported:

```{r output-options, eval=FALSE}
# Export only specific variables
write_gmpl(model, "model.mod", 
           include_solve = TRUE,
           export_vars = c("vTechCap", "vTechAct", "vObjective"))

# No custom CSV export (minimal file)
write_gmpl(model, "model.mod", 
           include_solve = TRUE,
           export_vars = FALSE)

# Use GMPL table output
write_gmpl(model, "model.mod", 
           include_solve = TRUE,
           use_table_output = TRUE)

# Combine: custom CSVs for key vars + table output for everything
write_gmpl(model, "model.mod", 
           include_solve = TRUE,
           export_vars = c("vObjective", "vTechCap"),
           use_table_output = TRUE)
```

### What Happens During write_gmpl_data()?

When `write_gmpl_data()` is called:

1. For each mapping/parameter, checks if data is in memory
2. If not, calls `get_lazy_data()` which:
   - Resolves the path: `base_path + misc$path`
   - Opens the Arrow/Parquet dataset
   - Collects data into a data frame
3. Writes the data in GMPL format with proper syntax:
   - Empty sets/mappings: `set name := ;`
   - Parameters with defaults: `param name default <defVal> := ...`
   - Comma-separated tuples: `elem1,elem2, elem3,elem4, ...`
   - Comma-separated indices: `[idx1,idx2] value`
4. Data is immediately eligible for garbage collection

This means:
- **Memory efficient**: Only one parameter in memory at a time
- **Transparent**: Same code works for in-memory or on-disk data
- **Scalable**: Can handle models with hundreds of parameters

## Step 5: Inspect Generated Files

Let's look at what was generated.

```{r inspect-mod}
# Show first 50 lines of .mod file
cat("\n=== Model file (.mod) - First 50 lines ===\n")
mod_lines <- readLines(mod_file, n = 50)
cat(paste(mod_lines, collapse = "\n"), "\n")
```

```{r inspect-dat}
# Show first 100 lines of .dat file
cat("\n=== Data file (.dat) - First 100 lines ===\n")
dat_lines <- readLines(dat_file, n = 100)
cat(paste(dat_lines, collapse = "\n"), "\n")
```

### Data File Structure

The `.dat` file contains:

```ampl
set FORIF := FORIFSET;  # Conditional evaluation set

# Basic sets (derived from data)
set region := R1 R2 R3 R4 R5 R6 R7;
set year := 2025 2026 2030 2033;
set comm := ELC COA GAS NUC HYD ...;

# Mappings (tuples)
set mCommReg :=
CO2 R1,
CO2 R2,
ELC R1,
...;

# Parameters (values with indices)
param pDemand :=
[DEM_ELC R1 2025] 100.5,
[DEM_ELC R2 2025] 85.3,
...;
```

## Step 6: Solve with GLPK

Now we solve the model using GLPK via the glpkAPI interface.

```{r solve}
cat("\n=== SOLVING WITH GLPK ===\n\n")

# Execute GLPK solver
result <- execute_glpkAPI(
  model_dir = output_dir,
  mod_file = "utopia.mod",
  dat_file = "utopia.dat",
  verbose = TRUE,
  save_solution = TRUE
)

# Display results
cat("\n=== SOLUTION SUMMARY ===\n")
cat("Status:", result$status_message, "\n")
cat("Objective value:", format(result$objective, big.mark = ",", scientific = FALSE), "\n")
cat("Variables:", result$n_variables, "\n")
cat("Constraints:", result$n_constraints, "\n")
cat("Non-zero variables:", nrow(result$solution_nonzero), "\n")
```

### Solution Output

The solver automatically executes the `printf` statements in the model file, writing solution CSV files to the `output/` directory:

- `output/log.csv` - Solver timing information
- `output/vObjective.csv` - Objective function value
- `output/vTechCap.csv` - Technology capacity variables
- `output/vBalance.csv` - Commodity balance variables
- ... (one file per variable)

```{r inspect-solution}
# Show objective value
cat("\n=== OBJECTIVE VALUE ===\n")
obj_file <- file.path(output_dir, "output", "vObjective.csv")
if (file.exists(obj_file)) {
  obj_data <- read.csv(obj_file)
  print(obj_data)
}

# Show non-zero variables summary
cat("\n=== NON-ZERO VARIABLES (first 20) ===\n")
print(head(result$solution_nonzero, 20))
```

## Performance Considerations

### Memory Usage

The lazy loading approach keeps memory usage low:

```{r memory-profile, eval=FALSE}
# Before write_gmpl_data()
mem_before <- pryr::mem_used()

# Write data file (loads from disk)
write_gmpl_data(model, file = dat_file)

# After (data eligible for GC)
mem_after <- pryr::mem_used()

cat("Memory increase:", format(mem_after - mem_before, units = "MB"), "\n")
```

### Timing

```{r timing, eval=FALSE}
# Time the complete workflow
system.time({
  model_struct <- read_gams(gams_file)
  model <- as_multimod(model_struct)
  # ... populate data ...
  write_gmpl(model, file = mod_file)
  write_gmpl_data(model, file = dat_file)
})

# Time the solve
system.time({
  result <- execute_glpkAPI(output_dir, verbose = FALSE)
})
```

## Comparison with Original Model

We can compare our solution with energyRt's original GLPK output if available:

```{r compare}
# Try to read original objective from output files
original_output_dir <- "C:/Users/admin/Documents/R/multimod/dev/scenarios/BASE_UTOPIA/script/glpk/output"
orig_obj_file <- file.path(original_output_dir, "vObjective.csv")

if (file.exists(orig_obj_file)) {
  orig_obj_data <- read.csv(orig_obj_file)
  original_obj <- orig_obj_data$value[1]
  
  # Our model objective
  our_obj <- result$objective
  
  # Difference
  diff <- abs(our_obj - original_obj)
  rel_diff <- diff / abs(original_obj) * 100
  
  cat("\nComparison with original:\n")
  cat("Original objective:", format(original_obj, scientific = FALSE), "\n")
  cat("Our objective:     ", format(our_obj, scientific = FALSE), "\n")
  cat("Absolute diff:     ", format(diff, scientific = TRUE), "\n")
  cat("Relative diff:     ", sprintf("%.2e%%", rel_diff), "\n")
  
  if (rel_diff < 1e-6) {
    cat("\n✓ Solutions match (numerically equivalent)\n")
  } else if (rel_diff < 1) {
    cat("\n✓ Solutions close (< 1% difference)\n")
  }
} else {
  cat("\nOriginal solution not found for comparison\n")
  cat("Our objective:", format(result$objective, scientific = FALSE), "\n")
}
```

### Compare CSV Outputs

Compare individual variable CSV files:

```{r compare-csvs}
# Paths to output directories
original_output_dir <- "C:/Users/admin/Documents/R/multimod/dev/scenarios/BASE_UTOPIA/script/glpk/output"
our_output_dir <- file.path(output_dir, "output")

if (dir.exists(original_output_dir)) {
  original_files <- list.files(original_output_dir, pattern = "\\.csv$")
  our_files <- list.files(our_output_dir, pattern = "\\.csv$")
  
  # Find common files
  common_files <- intersect(original_files, our_files)
  cat("\nComparing", length(common_files), "common CSV files:\n\n")
  
  for (fname in head(common_files, 10)) {
    orig_path <- file.path(original_output_dir, fname)
    our_path <- file.path(our_output_dir, fname)
    
    tryCatch({
      orig_data <- read.csv(orig_path)
      our_data <- read.csv(our_path)
      
      if (nrow(orig_data) == nrow(our_data) && "value" %in% names(orig_data)) {
        max_diff <- max(abs(orig_data$value - our_data$value), na.rm = TRUE)
        rel_diff <- max_diff / (max(abs(orig_data$value), na.rm = TRUE) + 1e-10)
        status <- if (rel_diff < 1e-6) "✓ MATCH" else sprintf("⚠ DIFF: %.2e", rel_diff)
      } else {
        status <- sprintf("%d rows", nrow(our_data))
      }
      
      cat(sprintf("  %-25s %s\n", fname, status))
    }, error = function(e) {
      cat(sprintf("  %-25s ERROR\n", fname))
    })
  }
}
```

## Complete Workflow Script

Here's the complete workflow in a single script:

```{r complete-workflow, eval=FALSE}
library(multimod)
library(energyRt)

# 1. Read and convert GAMS model
gams_file <- "path/to/energyRt.gms"
model <- as_multimod(read_gams(gams_file))

# 2. Load scenario and import data
load("path/to/scen.RData")
model$base_path <- "path/to/scenario"

# Import all data (sets, parameters, mappings, bounds)
model <- import_energyRt_data(model, scen, inMemory = TRUE)

# 3. Write GMPL files
output_dir <- "path/to/output"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

write_gmpl(model, 
           file = file.path(output_dir, "model.mod"), 
           include_solve = TRUE)

write_gmpl_data(model, 
                file = file.path(output_dir, "model.dat"))

# 4. Solve with GLPK
result <- execute_glpkAPI(output_dir, 
                         mod_file = "model.mod",
                         dat_file = "model.dat",
                         verbose = TRUE, 
                         save_solution = TRUE)

# 5. Analyze results
cat("Status:", result$status_message, "\n")
cat("Objective:", result$objective, "\n")
cat("Non-zero variables:", nrow(result$solution_nonzero), "\n")
print(head(result$solution_nonzero, 20))
```

## Diagnostics and Debugging

### Data Export for Inspection

When troubleshooting model generation, you can export the loaded sets and parameters to CSV format for inspection:

```{r export-data, eval=FALSE}
# Export all data to CSV files
write_gmpl(model, 
           file = "model.mod",
           export_data = TRUE)

# This creates a data/ directory with:
# - data/sets/*.csv       - One file per set
# - data/mappings/*.csv   - One file per mapping
# - data/parameters/*.csv - One file per parameter
```

Each CSV file contains the loaded data in a readable tabular format, making it easy to verify that data was imported correctly from energyRt scenarios.

### Solution Comparison

Compare solutions from different model runs or solvers:

```{r compare-solutions, eval=FALSE}
# Compare two solved models
result1 <- execute_glpkAPI("path/to/model1", verbose = FALSE)
result2 <- execute_glpkAPI("path/to/model2", verbose = FALSE)

comparison <- compare_solution(result1, result2)

# Returns:
# $objective_diff      - Difference in objective values
# $missing_in_first    - Variables present in second but not first
# $missing_in_second   - Variables present in first but not second
# $value_differences   - Variables with different values
```

You can also compare dumped CSV files from `export_data = TRUE`:

```{r compare-data, eval=FALSE}
# Compare parameter data between two models
compare_solution("model1/data/parameters/pDemand.csv",
                "model2/data/parameters/pDemand.csv")
```

### LP/MPS Export

Export models to standard LP or MPS formats for external validation or use with other solvers:

```{r export-lp-mps, eval=FALSE}
# Export to CPLEX LP format
export_glpk_model(model_dir = output_dir,
                 mod_file = "model.mod",
                 dat_file = "model.dat",
                 output_file = "model.lp",
                 format = "CPLEX_LP")

# Export to MPS format
export_glpk_model(model_dir = output_dir,
                 mod_file = "model.mod",
                 dat_file = "model.dat",
                 output_file = "model.mps",
                 format = "MPS")

# Also supports: "GLPK_LP" (GLPK native), "Fixed_MPS" (fixed-format MPS)
```

These exported files can be:
- Opened in text editors for manual inspection
- Used with other optimization solvers (CPLEX, Gurobi, CBC, etc.)
- Compared between model versions for debugging

### Deep Model Comparison

For detailed model comparison, use the LP/MPS comparison functions:

```{r compare-lp-mps, eval=FALSE}
# Compare LP files constraint-by-constraint and variable-by-variable
lp_comparison <- compare_lp("original.lp", "modified.lp", verbose = TRUE)

# Compare MPS files
mps_comparison <- compare_mps("original.mps", "modified.mps", verbose = TRUE)

# Both functions:
# - Extract all constraints and variables to data frames
# - Sort by name for order-independent comparison
# - Use compare::compare() for detailed difference reporting
# - Return rich comparison results

# Control comparison behavior:
compare_lp("model1.lp", "model2.lp",
          ignoreOrder = TRUE,        # Default: ignore row/column ordering
          ignoreNameCase = FALSE,    # Default: names are case-sensitive
          verbose = TRUE)            # Show progress during comparison
```

The comparison reports:
- **Constraints**: name, type (≤, ≥, =), lower/upper bounds
- **Variables**: name, type (continuous, integer, binary), bounds, objective coefficients
- **Differences**: Missing elements, value mismatches, structural changes

This is particularly useful for:
- **Roundtrip validation**: Verify that GAMS → multimod → GMPL preserves the model
- **Refactoring**: Confirm that code changes don't alter the mathematical model
- **Debugging**: Identify exactly what changed between model versions
- **Solver comparison**: Ensure different solvers receive equivalent models

### Complete Diagnostics Workflow

```{r diagnostics-workflow, eval=FALSE}
# 1. Export data for inspection
write_gmpl(model, "model.mod", export_data = TRUE)

# 2. Generate and solve model
result1 <- execute_glpkAPI(output_dir, verbose = TRUE)

# 3. Export to LP format for archival/comparison
export_glpk_model(output_dir, "model.mod", "model.dat", 
                 output_file = "baseline.lp", format = "CPLEX_LP")

# 4. Make changes to model...
# model_modified <- modify_model(model)
# write_gmpl(model_modified, "model.mod")
# write_gmpl_data(model_modified, "model.dat")

# 5. Solve modified model
result2 <- execute_glpkAPI(output_dir, verbose = TRUE)

# 6. Export modified version
export_glpk_model(output_dir, "model.mod", "model.dat",
                 output_file = "modified.lp", format = "CPLEX_LP")

# 7. Compare everything
compare_solution(result1, result2)  # Compare solutions
compare_lp("baseline.lp", "modified.lp", verbose = TRUE)  # Deep model comparison
```

## Summary

This workflow demonstrates:

- ✅ **Seamless conversion** from GAMS to GMPL
- ✅ **Lazy loading** from Arrow/Parquet datasets
- ✅ **Memory efficiency** for large models
- ✅ **Automated solving** with GLPK
- ✅ **Solution extraction** to CSV files
- ✅ **Comprehensive diagnostics** for debugging and validation

The key innovation is the lazy loading system, which allows working with large energyRt scenarios without loading all data into memory at once.

## Further Reading

- GMPL language reference: [GLPK documentation](https://www.gnu.org/software/glpk/)
- energyRt package: [GitHub repository](https://github.com/energyRt/energyRt)
- Arrow format: [Apache Arrow documentation](https://arrow.apache.org/)

## Session Info

```{r sessioninfo}
sessionInfo()
```
