---
title: "multimod to JuMP"
author: "multimod Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{multimod to JuMP}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE to run during build
)
```

## Introduction

This vignette demonstrates the complete workflow for converting an energyRt model to JuMP (Julia for Mathematical Programming) format and solving it with modern solvers like HiGHS, Gurobi, or CPLEX. The workflow includes:

1. Reading a GAMS model structure
2. Converting to multimod format
3. Connecting to energyRt scenario data
4. Writing JuMP model files
5. Solving with Julia/JuMP

We'll use the UTOPIA base scenario from energyRt as our example.

## JuMP-Specific Features

### Key Functions

#### 1. `write_jump(model, model_dir, cleanup, format)`

Generates a complete Julia/JuMP project with model code, data loading, and diagnostics.

**Parameters:**

- `model_dir`: Output directory for the project
- `cleanup = FALSE`: Whether to remove existing directory first
- `format = "ipc"`: Data format ("ipc" for Arrow, "csv" for CSV files)

**Generated Files:**

```
model_dir/
├── format.txt              # Data format specification
├── model.rds               # Saved R model object
├── sets/                   # Set members
├── mappings/               # Set membership data
├── parameters/             # Numeric parameter data
└── solvers/
    └── jump/
        ├── model.jl        # Main model file
        ├── data.jl         # Data loading code
        ├── constraint_stats.csv  # Diagnostic output
        └── variable_stats.csv    # Diagnostic output
```

**Key Features:**

- **Modern JuMP syntax**: Named constraints with tuple indexing
- **Short index aliases**: `h` (tech), `r` (region), `c` (comm), `y` (year), `ts` (slice)
- **get() pattern**: Clean parameter access with defaults
- **Variable bounds**: Properly typed (>=0, <=0, binary, integer, free)
- **Comprehensive diagnostics**: Constraint and variable statistics with nonzero counts

#### 2. `save_model(model, model_dir, format, verbose)`

Saves model data (sets, mappings, parameters) to disk in Arrow or CSV format.

**Parameters:**

- `format = "ipc"`: Arrow IPC format (fast, compact)
- `format = "csv"`: CSV format (human-readable, slower)
- `verbose = TRUE`: Print progress information

**Arrow Format Benefits:**

- **Fast**: Binary format, instant loading
- **Compact**: 40-60% smaller than CSV
- **Type-safe**: Preserves data types exactly
- **Memory-mapped**: Can load without copying to RAM

### Modern JuMP Code Generation

The generated JuMP code uses modern best practices:

```julia
# Named constraints with tuple indexing
eqTechCap = @constraint(
    model,
    eqTechCap[(h, y) in mTechSpan],
    vTechCap[(h, y)] == 
        get(pTechStock, (h, y), pTechStockDef) +
        sum(vTechNewCap[(h, yy)] for yy in year 
            if (h, yy, y) in mTechNew)
)

# Variables with proper bounds
@variable(model, vTechCap[mTechSpan] >= 0)
@variable(model, vBalance[mvBalance] >= -Inf)  # free variable
@variable(model, vTechRetired[mTechRetired], Bin)  # binary

# Clean parameter access with defaults
get(pTechStock, (h, y), pTechStockDef)

# Constraint diagnostics (automatic)
_log_constraint(constraint_stats, "eqTechCap", eqTechCap, 
                length(mTechSpan), time() - _t0)
```

### Diagnostics and Validation

Every generated model includes automatic diagnostics:

**Constraint Statistics:**
- Count: Number of constraint instances
- Dimension: Number of indices
- Mapping size: Size of constraint domain
- Time: Construction time in seconds
- Nonzeros: Total, min, max, average coefficients per constraint

**Variable Statistics:**
- Count: Number of variable instances
- Dimension: Number of indices
- Bounds: Fixed, free, lower_only, upper_only, bounded

Output CSV files enable:
- Model verification and comparison
- Performance profiling
- Debugging dimension mismatches
- Tracking constraint complexity

## Prerequisites

```{r libraries}
library(multimod)
library(energyRt)  # For scenario data
```

**Julia Requirements:**

- Julia 1.6+ (preferably 1.10+)
- JuMP.jl package
- Solver package (HiGHS.jl, Gurobi.jl, CPLEX.jl, etc.)
- Arrow.jl (for IPC format) or CSV.jl (for CSV format)
- DataFrames.jl

Install Julia packages:
```julia
using Pkg
Pkg.add(["JuMP", "HiGHS", "Arrow", "DataFrames", "CSV", "Dates"])
```

## Step 1: Read GAMS Model Structure

```{r read-gams}
# Path to GAMS model
gams_file <- "C:/Users/admin/Documents/R/multimod/dev/scenarios/BASE_UTOPIA/script/gams_cbc/energyRt.gms"

# Read model structure
cat("Reading GAMS model...\n")
model_struct <- read_gams(gams_file)

# Inspect structure
cat("Model structure:\n")
cat("  Sets:", length(model_struct$sets), "\n")
cat("  Mappings:", length(model_struct$mappings), "\n")
cat("  Parameters:", length(model_struct$parameters), "\n")
cat("  Variables:", length(model_struct$variables), "\n")
cat("  Equations:", length(model_struct$equations), "\n")
```

## Step 2: Convert to Multimod Format

```{r convert-multimod}
cat("\nConverting to multimod...\n")
model <- as_multimod(model_struct)

# Extract domain information from GAMS comments
model <- en_extract_domains_from_comments(model, verbose = TRUE)

# Populate default values from energyRt conventions
model <- populate_defvals_from_energyrt(model)

# Check index aliases (used in JuMP code)
cat("\nIndex aliases for JuMP:\n")
print(model$index_aliases)
```

The index aliases are used throughout the generated JuMP code:
```julia
for (h, r, c, y, ts) in mvTechInp  # not (tech, region, comm, year, slice)
    vTechInp[(h, r, c, y, ts)]      # tuple indexing
end
```

## Step 3: Load and Import Scenario Data

```{r load-scenario}
# Load energyRt scenario
cat("\nLoading energyRt scenario...\n")
scen_file <- "C:/Users/admin/Documents/R/multimod/dev/scenarios/BASE_UTOPIA/scen.RData"
load(scen_file)

cat("Scenario class:", class(scen), "\n")
```

### Import with Comprehensive Logging

```{r import-data}
# Import all data (sets, parameters, mappings, bounds)
cat("\nImporting energyRt scenario data...\n")
model <- import_energyRt_data(
  model, 
  scen, 
  inMemory = TRUE,  # Load data into R memory
  log_file = "tmp/import_log.csv"
)

cat("\nImport complete!\n")

# Verify data population
n_params <- length(model$parameters)
n_with_data <- sum(sapply(model$parameters, function(p) {
  !is.null(p$data) && nrow(p$data) > 0
}))

cat("Parameters with data:", n_with_data, "/", n_params, "\n")
```

### Check Bounds Processing

The importer automatically splits bounds parameters:

```{r check-bounds}
# Show bounds mapping
show_bounds_mapping(model)

# Example: pTechAf (type="bounds") becomes:
#   pTechAfLo - lower bounds
#   pTechAfUp - upper bounds
```

## Step 4: Save Model Data

Save the model data to disk for Julia to load:

```{r save-data}
# Create output directory
test_dir <- "tmp/utopia_jump"

# Save model data in Arrow IPC format (recommended)
cat("\nSaving model data (Arrow IPC format)...\n")
save_model(model, test_dir, format = "ipc", verbose = TRUE)

# Alternative: CSV format (slower, but human-readable)
# save_model(model, test_dir, format = "csv", verbose = TRUE)

cat("\nData saved to:", test_dir, "\n")
```

**Format Comparison:**

| Format | Speed | Size | Human-readable | Type-safe |
|--------|-------|------|----------------|-----------|
| IPC (Arrow) | ⚡⚡⚡ Fast | 📦 Compact | ❌ No | ✅ Yes |
| CSV | 🐌 Slow | 📦📦 Large | ✅ Yes | ⚠️ Partial |

## Step 5: Generate JuMP Code

```{r write-jump}
# Generate complete JuMP project
cat("\nGenerating JuMP code...\n")
write_jump(model, model_dir = test_dir)

cat("✓ JuMP code generated\n")

# Show generated files
julia_dir <- file.path(test_dir, "solvers", "jump")
cat("\nGenerated files:\n")
print(list.files(julia_dir, recursive = FALSE))
```

### Inspect Generated Code

```{r inspect-model}
# Show first 100 lines of model.jl
model_file <- file.path(julia_dir, "model.jl")
cat("\n=== model.jl (first 100 lines) ===\n")
model_lines <- readLines(model_file, n = 100)
cat(paste(model_lines, collapse = "\n"), "\n")
```

### Inspect Data Loading Code

```{r inspect-data}
# Show data loading code
data_file <- file.path(julia_dir, "data.jl")
cat("\n=== data.jl (first 80 lines) ===\n")
data_lines <- readLines(data_file, n = 80)
cat(paste(data_lines, collapse = "\n"), "\n")
```

Key features in data.jl:

```julia
# Automatic format detection
DATA_EXT = lowercase(strip(read("../../format.txt", String))) == "csv" ? ".csv" : ".arrow"

# Smart 1D vs multi-D handling
function _load_mapping_file(mapping_name)
    # ...
    if ncol(df) == 1
        return Set(string(row[1]) for row in eachrow(df))  # Set{String}
    else
        return Set(Tuple(...) for row in eachrow(df))       # Set{Tuple{...}}
    end
end

# Parameters with dimension-aware keys
function _load_param_file(param_name, dims)
    # ...
    if length(dims) == 1
        return Dict(string(row[1]) => row[end] ...)  # Scalar keys
    else
        return Dict(Tuple(...) => row[end] ...)       # Tuple keys
    end
end
```

This ensures:
- ✅ `t in mTradeCapacityVariable` works (not `(t,) in ...`)
- ✅ `get(pSliceWeight, "ANNUAL", 0.0)` works (not tuple)
- ✅ `get(pTechStock, (h, y), 0.0)` works with tuples

## Step 6: Solve with Julia/JuMP

### Option A: Run from R

```{r solve-from-r}
# Execute Julia model from R
cat("\n=== SOLVING WITH JULIA/JUMP ===\n\n")

# Change to Julia directory and run
old_wd <- getwd()
setwd(julia_dir)
system2("julia", args = "model.jl")
setwd(old_wd)
```

### Option B: Run from Julia REPL

```julia
# In Julia REPL:
cd("path/to/tmp/utopia_jump/solvers/jump")
include("model.jl")
```

### Option C: Run from Terminal

```bash
cd path/to/tmp/utopia_jump/solvers/jump
julia model.jl
```

## Step 7: Analyze Results

### Read Diagnostic CSVs

```{r read-diagnostics}
# Read constraint statistics
constraint_csv <- file.path(julia_dir, "constraint_stats.csv")
if (file.exists(constraint_csv)) {
  cat("\n=== CONSTRAINT STATISTICS ===\n")
  con_stats <- read.csv(constraint_csv)
  
  cat("Total constraints:", nrow(con_stats), "\n")
  cat("Non-empty constraints:", sum(con_stats$count > 0), "\n")
  cat("Total instances:", sum(con_stats$count), "\n")
  cat("Total nonzeros:", sum(con_stats$nnz_total), "\n\n")
  
  # Top 10 by count
  cat("Top 10 largest constraints:\n")
  top10 <- head(con_stats[order(-con_stats$count), 
                          c("name", "count", "dimension", "nnz_total", "nnz_avg")], 10)
  print(top10)
  
  # Top 10 by construction time
  cat("\nTop 10 slowest to build:\n")
  top10_time <- head(con_stats[order(-con_stats$time_seconds), 
                               c("name", "count", "time_seconds", "nnz_total")], 10)
  print(top10_time)
}
```

```{r read-variables}
# Read variable statistics
variable_csv <- file.path(julia_dir, "variable_stats.csv")
if (file.exists(variable_csv)) {
  cat("\n=== VARIABLE STATISTICS ===\n")
  var_stats <- read.csv(variable_csv)
  
  cat("Total variables:", nrow(var_stats), "\n")
  cat("Total instances:", sum(var_stats$count), "\n\n")
  
  cat("Bounds summary:\n")
  cat("  Fixed:      ", sum(var_stats$fixed), "\n")
  cat("  Free:       ", sum(var_stats$free), "\n")
  cat("  Lower only: ", sum(var_stats$lower_only), "\n")
  cat("  Upper only: ", sum(var_stats$upper_only), "\n")
  cat("  Bounded:    ", sum(var_stats$bounded), "\n\n")
  
  # Top 10 by count
  cat("Top 10 largest variables:\n")
  top10 <- head(var_stats[order(-var_stats$count), 
                          c("name", "count", "dimension", "lower_only", "free")], 10)
  print(top10)
}
```

### Parse Solution from Julia Output

The Julia script prints solution information to console. You can capture it:

```{r parse-solution}
# Run and capture output
output <- system2("julia", 
                  args = file.path(julia_dir, "model.jl"),
                  stdout = TRUE, 
                  stderr = TRUE)

# Parse objective value
obj_line <- grep("Objective value:", output, value = TRUE)
if (length(obj_line) > 0) {
  obj_value <- as.numeric(sub(".*Objective value:\\s+([-0-9.e+]+).*", "\\1", obj_line))
  cat("\nObjective value:", format(obj_value, big.mark = ",", scientific = FALSE), "\n")
}

# Parse termination status
status_line <- grep("Termination status:", output, value = TRUE)
if (length(status_line) > 0) {
  status <- sub(".*Termination status:\\s+(\\w+).*", "\\1", status_line)
  cat("Termination status:", status, "\n")
}

# Parse solve time
time_line <- grep("HiGHS run time", output, value = TRUE)
if (length(time_line) > 0) {
  solve_time <- sub(".*HiGHS run time\\s*:\\s*([-0-9.]+).*", "\\1", time_line)
  cat("Solve time:", solve_time, "seconds\n")
}
```

## Comparison with energyRt Original

Compare with the original energyRt translation if available:

```{r compare-models}
# Load reference model results
ref_con_stats <- "dev/scenarios/BASE_UTOPIA/script/julia_highs/constraint_stats.csv"
if (file.exists(ref_con_stats)) {
  ref_con <- read.csv(ref_con_stats)
  our_con <- read.csv(constraint_csv)
  
  cat("\n=== MODEL COMPARISON ===\n\n")
  
  # Compare sizes
  cat("Constraint counts:\n")
  cat("  Reference: ", nrow(ref_con), "constraint families\n")
  cat("  Our model: ", nrow(our_con), "constraint families\n\n")
  
  cat("Total instances:\n")
  cat("  Reference: ", sum(ref_con$count), "\n")
  cat("  Our model: ", sum(our_con$count), "\n\n")
  
  cat("Total nonzeros:\n")
  cat("  Reference: ", sum(ref_con$nnz_total), "\n")
  cat("  Our model: ", sum(our_con$nnz_total), "\n\n")
  
  # Compare specific constraints
  common_names <- intersect(ref_con$name, our_con$name)
  cat("Comparing", length(common_names), "common constraints:\n\n")
  
  for (cname in head(common_names, 10)) {
    ref_row <- ref_con[ref_con$name == cname, ]
    our_row <- our_con[our_con$name == cname, ]
    
    match <- (ref_row$count == our_row$count && 
              ref_row$nnz_total == our_row$nnz_total)
    
    status <- if (match) "✓" else "✗"
    
    cat(sprintf("  %s %-20s: ref=%4d/%5d nnz, our=%4d/%5d nnz\n",
                status, cname, 
                ref_row$count, ref_row$nnz_total,
                our_row$count, our_row$nnz_total))
  }
}
```

## Advanced: Custom Solver Configuration

You can modify the generated `model.jl` to use different solvers or set solver options:

```julia
# Use Gurobi instead of HiGHS
using Gurobi
model = Model(Gurobi.Optimizer)
set_optimizer_attribute(model, "TimeLimit", 3600)
set_optimizer_attribute(model, "MIPGap", 1e-4)

# Use CPLEX
using CPLEX
model = Model(CPLEX.Optimizer)
set_optimizer_attribute(model, "CPX_PARAM_TILIM", 3600)

# Use Cbc (open-source)
using Cbc
model = Model(Cbc.Optimizer)
set_optimizer_attribute(model, "seconds", 3600)

# Multiple solvers with fallback
try
    using Gurobi
    model = Model(Gurobi.Optimizer)
catch
    using HiGHS
    model = Model(HiGHS.Optimizer)
end
```

## Performance Tips

### 1. Use Arrow IPC Format

Arrow is 3-5x faster than CSV for loading:

```{r benchmark-formats, eval=FALSE}
# Benchmark data loading
library(microbenchmark)

microbenchmark(
  arrow = {
    save_model(model, "tmp/bench_arrow", format = "ipc", verbose = FALSE)
  },
  csv = {
    save_model(model, "tmp/bench_csv", format = "csv", verbose = FALSE)
  },
  times = 5
)
```

### 2. Pre-compile Julia Package

First run is slow due to compilation. Pre-compile:

```julia
using Pkg
Pkg.precompile()
```

Or use PackageCompiler.jl to create a sysimage:

```julia
using PackageCompiler
create_sysimage(["JuMP", "HiGHS", "Arrow", "DataFrames"]; 
                sysimage_path="jump_sysimage.so")
```

Then run with: `julia --sysimage jump_sysimage.so model.jl`

### 3. Monitor Memory Usage

For large models, track memory:

```julia
using Logging
@info "Memory before model build" memory=Base.gc_live_bytes()/1e9

# Build model...

@info "Memory after model build" memory=Base.gc_live_bytes()/1e9
```

## Debugging Tips

### Check Data Loading

If the model fails to load, test data loading separately:

```julia
# In Julia REPL:
include("data.jl")

# Check specific sets/parameters
println("comm set: ", comm)
println("pTechStock: ", typeof(pTechStock), " with ", length(pTechStock), " entries")

# Test parameter access
println("Test get: ", get(pTechStock, ("E01", "2025"), 0.0))
```

### Check Constraint Dimensions

If constraints have unexpected nonzero counts:

```{r check-dimensions}
# In R, before generating code:
cat("\nChecking constraint dimensions:\n")
for (eq in model$equations[1:5]) {
  cat(sprintf("%-20s: %s\n", eq$name, paste(eq$dims, collapse = " x ")))
}

# Check mapping sizes
cat("\nChecking mapping sizes:\n")
for (map_name in names(model$parameters)[1:10]) {
  p <- model$parameters[[map_name]]
  if (!is.null(p$data)) {
    cat(sprintf("%-25s: %5d rows\n", map_name, nrow(p$data)))
  }
}
```

### Validate Against GAMS

Compare with GAMS output if available:

```{r compare-gams, eval=FALSE}
# Read GAMS solution
gams_gdx <- "path/to/gams_solution.gdx"
# ... use gdxrrw or other tools to read GDX
# Compare with Julia output
```

## Complete Workflow Script

Here's the complete end-to-end workflow:

```{r complete-workflow, eval=FALSE}
library(multimod)
library(energyRt)

# 1. Read and convert GAMS model
model <- as_multimod(read_gams("path/to/energyRt.gms"))
model <- en_extract_domains_from_comments(model)
model <- populate_defvals_from_energyrt(model)

# 2. Load and import scenario data
load("path/to/scen.RData")
model <- import_energyRt_data(model, scen, inMemory = TRUE)

# 3. Save model data
test_dir <- "tmp/my_model"
save_model(model, test_dir, format = "ipc", verbose = TRUE)

# 4. Generate JuMP code
write_jump(model, model_dir = test_dir)

# 5. Solve with Julia
julia_dir <- file.path(test_dir, "solvers", "jump")
setwd(julia_dir)
system2("julia", args = "model.jl")

# 6. Read results
con_stats <- read.csv(file.path(julia_dir, "constraint_stats.csv"))
var_stats <- read.csv(file.path(julia_dir, "variable_stats.csv"))

cat("Total constraints:", sum(con_stats$count), "\n")
cat("Total variables:", sum(var_stats$count), "\n")
cat("Total nonzeros:", sum(con_stats$nnz_total), "\n")
```

## Key Differences from GMPL

| Feature | GMPL | JuMP |
|---------|------|------|
| Language | AMPL-like | Julia (native) |
| Syntax | MathProg | Julia/JuMP macros |
| Data format | .dat files | Arrow/CSV |
| Index syntax | `[i,j,k]` | `[(i,j,k)]` tuple |
| Constraints | Anonymous | Named objects |
| Solvers | GLPK only | HiGHS, Gurobi, CPLEX, Cbc, ... |
| Performance | Good | Excellent |
| Ecosystem | Limited | Rich (Julia packages) |
| Debugging | Printf | Full Julia tooling |

## Summary

The JuMP translation provides:

- ✅ **Modern syntax**: Named constraints, tuple indexing, get() pattern
- ✅ **Type safety**: Proper handling of 1D vs multi-D data
- ✅ **Variable bounds**: From GAMS variable types
- ✅ **Diagnostics**: Comprehensive constraint/variable statistics
- ✅ **Performance**: Fast Arrow format, efficient code generation
- ✅ **Flexibility**: Multiple solver options
- ✅ **Validation**: Built-in nonzero counts for debugging

The key innovation is the dimension-aware data loading that correctly handles both scalar and tuple keys, ensuring generated code matches the original model structure exactly.

## Further Reading

- JuMP documentation: [jump.dev](https://jump.dev/)
- Julia language: [julialang.org](https://julialang.org/)
- Arrow format: [arrow.apache.org](https://arrow.apache.org/)
- energyRt package: [github.com/energyRt/energyRt](https://github.com/energyRt/energyRt)
- HiGHS solver: [highs.dev](https://highs.dev/)

## Session Info

```{r sessioninfo}
sessionInfo()
```
