---
title: "LaTeX Generation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LaTeX Generation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

This vignette demonstrates how to generate LaTeX documents from multimod models, including options for displaying folded and trimmed models with proper annotations.

## Setup

```{r setup}
library(multimod)
```

## Load Example Model

We'll use the UTOPIA model that comes with the multimod package:

```{r load-model}
data(example_models)
demo_model <- example_models$energyRt$multimod
print(demo_model)
```

## Basic LaTeX Generation

The simplest way to generate LaTeX is to call `write_latex()` on a model:

```{r basic-latex, eval=FALSE}
# Generate LaTeX as a string
tex <- write_latex(demo_model)

# Save to file
write_latex(demo_model, file = "model.tex")
```

The generated LaTeX document includes:
- Document preamble with necessary packages
- Title and metadata
- Sets with element listings
- Parameters with dimensions
- Variables with dimensions
- Equations with mathematical notation
- Mappings (domain definitions)

## Model View Modes

The `model_view` parameter controls how the model is displayed:

### Auto Mode (Default)
Automatically detects if model is optimized and chooses appropriate view:
```{r auto-mode, eval=FALSE}
tex <- write_latex(demo_model)  # Auto-detects view mode
```

### Reduced Mode
Shows only active (non-trimmed) elements, ideal for optimized models:
```{r reduced-mode, eval=FALSE}
tex <- write_latex(model_optimized, model_view = "reduced")
```

### Full Mode
Shows all elements with annotations for trimmed/folded items:
```{r full-mode, eval=FALSE}
tex <- write_latex(model_optimized, model_view = "full")
```

## Model Optimization and LaTeX Output

### Folding

Folding reduces parameter dimensionality by consolidating dimensions with uniform values:

```{r folding}
# Create fold specification
fold_spec <- create_fold_spec(
  demo_model,
  fold_dims = list(
    slice = list(
      tech = "mTechSlice",
      comm = "mCommSlice"
    ),
    region = list(
      tech = "mTechRegion",
      sup = "mSupRegion"
    )
  )
)

# Apply folding
model_folded <- fold_model(demo_model, fold_spec)

# Check folding summary
get_fold_summary(model_folded, format = "text")
```

### Trimming

Trimming removes empty and unused model elements:

```{r trimming}
# Trim the folded model
model_optimized <- trim_model(model_folded)

# Check trimming summary
get_trim_summary(model_optimized, format = "text")
```

### LaTeX with Optimization Summary

When generating LaTeX for optimized models, an optimization summary is automatically included:

```{r optimized-latex, eval=FALSE}
tex <- write_latex(model_optimized, model_view = "full")
```

The optimization summary includes:
- **Folding statistics**: Number and percentage of folded parameters
- **Dimension reductions**: Details of each folded parameter (e.g., 5D → 2D)
- **Trimming statistics**: Counts for trimmed sets, parameters, mappings, variables, equations
- **Explanatory text**: Describes what folding and trimming do

### Folded Annotations

In full view mode, folded parameters show green annotations:

```{r folded-annotations, eval=FALSE}
# Parameters will show: (folded: 5D → 2D) in green
tex <- write_latex(model_optimized, model_view = "full")
```

### Trimmed Annotations

Trimmed elements are marked in blue (customizable):

```{r trimmed-annotations, eval=FALSE}
# Default blue color
tex <- write_latex(model_optimized, 
                  model_view = "full", 
                  trimmed_color = "blue")

# Custom color
tex <- write_latex(model_optimized, 
                  model_view = "full", 
                  trimmed_color = "red")
```

## Data Display Options

### Include Data

Control whether to show set elements and parameter row counts:

```{r data-display, eval=FALSE}
# With data (default)
tex <- write_latex(demo_model, include_data = TRUE)

# Without data (theory mode)
tex <- write_latex(demo_model, include_data = FALSE)
```

### Data Detail Level

Control the amount of data shown:

```{r data-detail, eval=FALSE}
# Brief: Shows row counts and first few elements
tex <- write_latex(demo_model, data_detail = "brief")

# Detailed: Shows more information
tex <- write_latex(demo_model, data_detail = "detailed")
```

### Set Element Display

Customize how set elements are displayed:

```{r set-display, eval=FALSE}
tex <- write_latex(demo_model,
                  set_display_max_inline = 10,        # Max elements to show inline
                  set_display_head_tail_threshold = 100,  # Threshold for head/tail
                  set_display_head_n = 5,             # Number of head elements
                  set_display_tail_n = 3)             # Number of tail elements
```

## Index Aliases

Index aliases provide short names for dimensions in equations:

```{r aliases, eval=FALSE}
# Aliases are auto-detected from model$index_aliases
tex <- write_latex(demo_model, use_model_aliases = TRUE)

# This will use 'c' instead of 'comm', 'r' instead of 'region', etc.
```

The LaTeX output includes an "Index Aliases" section showing the mapping.

## Controlling Alias Usage

Control where aliases appear:

```{r alias-control, eval=FALSE}
# Default: Full names in declarations, aliases in equations
tex <- write_latex(demo_model, use_aliases_in_declarations = FALSE)

# Aliases everywhere
tex <- write_latex(demo_model, use_aliases_in_declarations = TRUE)
```

## Section Visibility

Control which sections appear in the output:

```{r sections, eval=FALSE}
tex <- write_latex(demo_model,
                  include_sets = TRUE,
                  include_aliases = TRUE,
                  include_index_aliases = TRUE,
                  include_parameters = TRUE,
                  include_variables = TRUE,
                  include_equations = TRUE,
                  include_mappings = TRUE)  # Now TRUE by default

# Equations only
tex <- write_latex(demo_model,
                  include_sets = FALSE,
                  include_parameters = FALSE,
                  include_variables = FALSE,
                  include_equations = TRUE)
```

## Table of Contents

Add a table of contents to the document:

```{r toc, eval=FALSE}
tex <- write_latex(demo_model, include_toc = TRUE)
```

## Complete Example

Here's a complete example generating a publication-ready LaTeX document:

```{r complete-example, eval=FALSE}
# Load and optimize model
data(example_models)
demo_model <- example_models$energyRt$multimod

# Create comprehensive fold specification
fold_spec <- create_fold_spec(
  demo_model,
  fold_dims = list(
    slice = list(
      tech = "mTechSlice",
      comm = "mCommSlice",
      stg = c("mStorageComm", "mCommSlice")
    ),
    region = list(
      tech = "mTechRegion",
      sup = "mSupRegion"
    ),
    year = list(
      tech = "mTechYear",
      stg = "mStorageYear"
    )
  )
)

# Apply folding and trimming
model_folded <- fold_model(demo_model, fold_spec)
model_optimized <- trim_model(model_folded)

# Generate comprehensive LaTeX document
write_latex(
  model_optimized,
  file = "utopia_optimized.tex",
  model_view = "full",              # Show all with annotations
  include_toc = TRUE,               # Include table of contents
  include_data = TRUE,              # Show data details
  data_detail = "brief",            # Brief data format
  use_model_aliases = TRUE,         # Use short names in equations
  use_aliases_in_declarations = FALSE,  # Full names in declarations
  trimmed_color = "blue",           # Color for trimmed annotations
  include_sets = TRUE,
  include_aliases = TRUE,
  include_index_aliases = TRUE,
  include_parameters = TRUE,
  include_variables = TRUE,
  include_equations = TRUE,
  include_mappings = TRUE
)

# Compile with pdflatex
system("pdflatex utopia_optimized.tex")
```

## Underscore Handling

The LaTeX generator automatically escapes underscores in set element names, ensuring they display correctly:

```{r underscores, eval=FALSE}
# Set elements like "A_D", "RES_HYD" are automatically escaped
# They will appear correctly in the PDF as A_D, RES_HYD
tex <- write_latex(demo_model, include_data = TRUE)
```

## Converting Individual Objects

You can also convert individual model components to LaTeX:

```{r individual-objects, eval=FALSE}
# Convert a parameter
param_latex <- as_latex(demo_model$parameters$pTechCinp2use)

# Convert a variable
var_latex <- as_latex(demo_model$variables$vTechInp)

# Convert an equation
eq_latex <- as_latex(demo_model$equations$eqTechInp2Out)

# Convert an AST node
ast_latex <- as_latex(demo_model$equations$eqTechInp2Out$rhs)
```

## Tips for Publication-Quality Output

1. **Use full mode with optimized models** to show readers what was removed
2. **Include table of contents** for longer models
3. **Add explanatory text** via the optimization summary
4. **Use consistent colors** for annotations (blue for trimmed, green for folded)
5. **Show brief data** to give readers context without overwhelming detail
6. **Use index aliases** to make equations more readable
7. **Keep full names in declarations** for clarity

## Compiling LaTeX Documents

After generating the `.tex` file, compile it with:

```bash
pdflatex model.tex
```

For models with table of contents, run twice:

```bash
pdflatex model.tex
pdflatex model.tex
```

## Summary

The `write_latex()` function provides comprehensive control over LaTeX generation:

- **Model views**: Auto, reduced, full, or both
- **Data display**: Control detail level and element formatting
- **Optimization**: Automatic summary with explanatory text
- **Annotations**: Visual markers for folded and trimmed elements
- **Aliases**: Short names for readability
- **Sections**: Fine-grained control over content
- **TOC**: Optional table of contents
- **Customization**: Colors, fonts, and display options

This makes it easy to generate publication-ready mathematical documentation directly from your model code.
