---
title: "Model Workspace and Data Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Workspace and Data Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(multimod)
```

## Overview

The `multimod` package provides a flexible workspace system for managing models and their data on disk. This vignette demonstrates:

- Creating and managing model workspaces
- Saving and loading models in different formats
- Working with in-memory vs on-disk data
- Updating parameters and mappings
- Lazy loading for large models
- Cross-language data interoperability

## Storage Architecture

### Hybrid Approach

The storage system uses a **hybrid approach** that balances R-specific needs with cross-language compatibility:

- **model.rds**: Complete R model object (structure, metadata, equations, variables, ASTs)
- **Data files**: Stored separately in cross-language formats (Arrow IPC, CSV, Parquet)
- **metadata.json**: Human-readable summary for non-R users

This design allows:
- Fast loading of R model structure (single RDS file)
- Cross-language access to data (Python, Julia can read Arrow/CSV)
- Atomic updates (one model.rds contains all structure)

### Directory Structure

```
model_path/
  model.rds              # Complete R model object with all structure
  metadata.json          # Human-readable summary
  format.txt             # Data format: "csv", "ipc", or "parquet"
  solution_log.txt       # Log of solution loads (appended on each load)
  sets/
    set_name.txt         # Set members (one per line)
  parameters/
    param_name/
      data.{csv|arrow|parquet}      # Parameter data only
      folded_data/
        data.{csv|arrow|parquet}    # Folded data if exists
  mappings/
    mapping_name/
      data.{csv|arrow|parquet}      # Mapping data only
  variables/             # Solution values (moved from solver on load_results)
    var_name/
      data.{csv|arrow|parquet}      # Variable solution values
  solvers/
    jump/                # JuMP (Julia) solver files
      model.jl           # Generated Julia/JuMP code
      data.jl            # Data loading functions
      solution/          # Latest solve results (overwritten each run)
        var_name/
          data.{csv|arrow|parquet}  # Variable solution values
    gmpl/                # GMPL (GLPK) solver files
      model.mod          # Generated GMPL code
```

## Storage Formats

### CSV (Universal)

```{r csv-format}
# Default format - human-readable, works everywhere
model <- save_model(
  model,
  path = "models/my_model",
  format = "csv"
)
```

**Pros:**
- Universal compatibility
- Human-readable
- Easy debugging

**Cons:**
- Largest file size
- Slower I/O

### IPC/Arrow (Recommended)

```{r ipc-format}
# Recommended format - fast and compact
model <- save_model(
  model,
  path = "models/my_model",
  format = "ipc",
  compression = "zstd",
  compression_level = 15
)
```

**Pros:**
- Fast I/O (10-20x faster than CSV)
- Good compression (2-3x smaller than CSV)
- Cross-language (Python, Julia, R)
- Column-oriented (efficient queries)

**Cons:**
- Binary format (not human-readable)
- Requires Arrow library

### Parquet (Compact)

```{r parquet-format}
# Most compact but slower
model <- save_model(
  model,
  path = "models/my_model",
  format = "parquet",
  compression = "zstd",
  compression_level = 15
)
```

**Pros:**
- Best compression (3-4x smaller than CSV)
- Wide ecosystem support

**Cons:**
- Slower than IPC
- More complex format

## Basic Workflow

### 1. Save a Model

```{r save-basic}
library(multimod)

# Load or create a model
data(example_models)
demo_model <- example_models$energyRt$multimod
model <- demo_model

# Save with default settings (CSV format)
model <- save_model(
  model,
  path = "models/utopia",
  verbose = TRUE
)
```

### 2. Load a Model

```{r load-basic}
# Load complete model with all data
model <- load_model(
  path = "models/utopia",
  verbose = TRUE
)

# Access parameters
tech_cost <- model$parameters$pTechCost$data
demand <- model$parameters$pDemand$data
```

### 3. Update Parameters

```{r update-basic}
# Modify parameter data
new_demand <- demand
new_demand$value <- new_demand$value * 1.1  # 10% increase

# Update in model
model <- update_parameter(
  model,
  name = "pDemand",
  data = new_demand
)

# Save changes
model <- save_model(model, path = "models/utopia", overwrite = TRUE)
```

## Advanced Features

### Lazy Loading

For large models, load structure without data:

```{r lazy-loading}
# Load only model structure (no data in memory)
model <- load_model(
  path = "models/utopia",
  load_data = FALSE
)

# Model structure is available
print(names(model$parameters))
print(model$parameters$pDemand$sets)

# Load specific parameters on demand
demand_data <- get_data(model, "pDemand", type = "parameter")

# get_data() automatically loads from disk and caches in model
# Subsequent calls use the cached data
demand_data2 <- get_data(model, "pDemand", type = "parameter")  # From cache
```

### Auto-Save on Update

```{r auto-save}
# Load model from disk
model <- load_model("models/utopia")

# Update parameter with immediate save to disk
model <- update_parameter(
  model,
  name = "pDemand",
  data = new_demand_data,
  auto_save = TRUE  # Writes immediately to disk
)

# No need for separate save_model() call
```

### Memory Management

```{r memory-management}
# Save model and free data from memory
model <- save_model(
  model,
  path = "models/utopia",
  keep_in_memory = FALSE  # Removes data from model object
)

# Model structure remains, but data is NULL
is.null(model$parameters$pDemand$data)  # TRUE

# Use get_data() to load from disk when needed
demand <- get_data(model, "pDemand", type = "parameter")
```

### Solution Management

The package provides integrated solution storage for solver results:

```{r solution-management}
# Solve model and load results
result <- solve_jump(
  model = demo_model,
  model_dir = "models/utopia",
  load_results = TRUE
)

# Update model with loaded solutions
model <- result$model

# Solutions are now in model$variables[[var_name]]$solution
vObjective <- model$variables$vObjective$solution
vTechCap <- model$variables$vTechCap$solution

# Solution metadata
print(model$solution_metadata$loaded_at)
print(model$solution_metadata$variables_loaded)

# Check solution log
log <- readLines("models/utopia/solution_log.txt")
print(tail(log, 5))
```

**Solution Storage Workflow:**

1. **Solver writes**: Solutions saved to `solvers/jump/solution/` (overwrites previous)
2. **Load results**: If `load_results=TRUE`, solutions copied to `variables/` (overwrites previous)
3. **Log append**: Each load appends timestamp and variable list to `solution_log.txt`

**Benefits:**
- Simple structure: one current solution in solver directory
- Audit trail: log tracks all solution loads with timestamps
- Clean workspace: no proliferation of timestamped directories
- Overwrite behavior: easy to re-run and update solutions
```

## Working with Folded Data

Models with folded parameters store both regular and folded data:

```{r folded-data}
# Apply folding
model <- fold_parameter(model, "pDemand", dimensions = "slice")

# Save model - both data and folded_data are saved
model <- save_model(model, path = "models/utopia_folded", format = "ipc")

# Load and access folded data
model <- load_model("models/utopia_folded")
demand_regular <- model$parameters$pDemand$data
demand_folded <- model$parameters$pDemand$folded_data$data

# Update with folded data
model <- update_parameter(
  model,
  name = "pDemand",
  data = new_regular_data,
  folded_data = new_folded_data
)
```

## Cross-Language Interoperability

### From R

```{r r-access}
# Save in Arrow IPC format
save_model(model, "models/shared", format = "ipc")
```

### From Python

```python
import pyarrow.feather as feather
import pandas as pd

# Read parameter data
demand = feather.read_feather("models/shared/parameters/pDemand/data.arrow")
print(demand.head())

# Read set members
with open("models/shared/sets/region.txt", "r") as f:
    regions = [line.strip() for line in f]
```

### From Julia

```julia
using Arrow

# Read parameter data
demand = Arrow.Table("models/shared/parameters/pDemand/data.arrow")
df = DataFrame(demand)
```

## Complete Example

```{r complete-example}
library(multimod)

# 1. Create or load model
data(example_models)
demo_model <- example_models$energyRt$multimod
model <- demo_model

# 2. Modify some parameters
demand <- model$parameters$pDemand$data
demand$value <- demand$value * 1.05  # 5% increase

model <- update_parameter(model, "pDemand", demand)

# 3. Save with recommended format
model <- save_model(
  model,
  path = "models/utopia_scenario1",
  format = "ipc",
  compression = "zstd",
  compression_level = 15,
  overwrite = TRUE,
  verbose = TRUE
)

# 4. Generate and solve with JuMP
write_jump(model, model_dir = "models/utopia_scenario1")

result <- solve_jump(
  model = model,
  model_dir = "models/utopia_scenario1",
  method = "system",
  load_results = TRUE
)

# Update model with solutions
model <- result$model

# 5. Access solution values
vObjective <- model$variables$vObjective$solution
vTechCap <- model$variables$vTechCap$solution

print(paste("Objective value:", vObjective$value))
print(head(vTechCap))

# 6. Check solution log
solution_log <- readLines("models/utopia_scenario1/solution_log.txt")
cat(tail(solution_log, 3), sep = "\n")

# 7. For large models: lazy loading
model_structure <- load_model(
  "models/utopia_scenario1",
  load_data = FALSE
)

# Load only needed parameters
demand <- get_data(model_structure, "pDemand", type = "parameter")
tech_cost <- get_data(model_structure, "pTechCost", type = "parameter")
```

## Performance Comparison

Approximate performance for a typical model:

| Format  | Size    | Save Time | Load Time | Compression |
|---------|---------|-----------|-----------|-------------|
| CSV     | 1.2 MB  | 100%      | 100%      | 1.0x        |
| IPC     | 0.6 MB  | 15%       | 10%       | 2.0x        |
| Parquet | 0.3 MB  | 40%       | 30%       | 3.6x        |

**Recommendation:** Use IPC format for best balance of speed, size, and compatibility.

## Best Practices

1. **Use IPC format** for production models (fast, compact, cross-language)
2. **Use lazy loading** for large models to minimize memory usage
3. **Use auto_save** when making incremental updates to on-disk models
4. **Keep model.rds** together with data files (don't separate them)
5. **Use version control** for model.rds (it's deterministic for same model state)
6. **Document changes** in metadata.json (edit manually if needed)
7. **Check solution_log.txt** to track when solutions were loaded and updated
8. **Use `load_results=TRUE`** to automatically move solutions from solver to model root

## Workflow Summary

```{r workflow-diagram, echo=FALSE, eval=FALSE}
# Complete workflow with solving:

# Import/Create Data
#       ↓
# update_parameter() ← Modify values
#       ↓
# save_model() ← Persist to disk
#       ↓
# write_jump() / write_gmpl() ← Generate solver code
#       ↓
# solve_jump() ← Solve and load results
#       ↓
# Access solutions in model$variables[[name]]$solution
#       ↓
# Check solution_log.txt for audit trail
```

**Import/Create** → **Update** → **Save** → **Generate** → **Solve** → **Access Solutions**

## Troubleshooting

### Large Models

```{r large-models}
# Use lazy loading
model <- load_model("models/large_model", load_data = FALSE)

# Process parameters one at a time
for (param_name in names(model$parameters)) {
  data <- get_data(model, param_name, type = "parameter")
  # ... process data ...
  # Clear from memory if needed
  model$parameters[[param_name]]$data <- NULL
}
```

### Disk Space

```{r disk-space}
# Use Parquet for maximum compression
save_model(model, "models/compact", format = "parquet")

# Or remove data from memory after saving
model <- save_model(model, "models/compact", keep_in_memory = FALSE)
```

### Corrupted Files

```{r recovery}
# Model structure is separate from data
# If data file corrupted, structure is still in model.rds

model <- readRDS("models/my_model/model.rds")

# Re-import or regenerate the corrupted parameter
model <- update_parameter(model, "pDemand", new_data, auto_save = TRUE)
```

## See Also

- `?save_model` - Save model to disk
- `?load_model` - Load model from disk
- `?update_parameter` - Update parameter data
- `?update_mapping` - Update mapping data
- `?get_data` - Lazy load data from disk
- `?solve_jump` - Solve model with JuMP and load results
- `?write_jump` - Generate JuMP/Julia code
- `?write_gmpl` - Generate GMPL code
- `vignette("jump")` - Complete JuMP workflow guide
- `vignette("gmpl")` - Complete GMPL workflow guide
- `vignette("folding")` - Parameter folding for model reduction
